The Talking Heads experiment

### 2015-04-28 chap2/chap2.tex
format relevant papers section

Virtual agents cannot interact and 
an interaction between two ``real'' agents
can only take place when they are both physically present
in the same location. Thus an agent can travel from
Paris to Tokyo at the blink of an eye, rather than having
to take a plane, but two agents can only interact when they
are instantiated in the same physical environment. 
It is in principle possible that agents in different 
physical locations describe to each other the environment that they 
see (but which the other one does not see), just as we 
would do in a telephone conversation. This is only possible,
however, after the agents have had sufficient interactions
with each other in a shared physical world to have 
developed and learned a grounded shared language.

@@

The speaker first looks at one area of the white board and 
directs the attention of the hearer to the same area.\footnote{I will explain later how exactly agents
decide on a particular scene and how they are able to 
draw each other's attention to specific areas of the 
white board.}
The objects located in this area constitute the {\itshape context}. 
The speaker then chooses one object from the context, which 
I will call the {\itshape topic}, and gives a verbal hint
to the hearer. The {\itshape verbal hint} is an expression
that identifies the topic 
with respect to the other objects in the context. For example, 
if the context contains a red square, a blue triangle, 
and a green circle, then the speaker may say something like 
`the red one' to identify the red square as the topic. 
If the context contains also a red triangle, he has to be more 
precise and say something like `the red square' to delineate
the topic from the red triangle as well as
from the blue square. 
Of course, the Talking Heads do not say `the red square'
but use their own language and concepts which are not 
necessarily the same as those used in English. For example, 
they may say `malewina' to mean [\textsc{upper extreme}-\textsc{left low}-\textsc{redness}].

Based on the verbal hint, the hearer tries to guess what
topic the speaker has chosen, and communicates his choice 
to the speaker by pointing to the object. Given that 
the robots do not have arms, pointing is realised by 
focusing on an object. One robot can ``see'' in which direction another one 
is looking, and thus know where he is ``pointing''. 
The game succeeds if the topic guessed by the hearer is 
equal to the topic chosen by the speaker. 
The game fails if the guess was wrong or 
if the speaker or the hearer failed at some earlier point in the 
game.

@@

The Guessing Game is one of the common things we do with language. 
For example, I play a similar game when I sit with a friend
at the dining table and say `could you give me the salt'. 
If she guesses correctly what I mean and hands 
me the salt, the game has succeeded. If she looks at me 
with a puzzled face (maybe she does not speak English) or hands me
the salmon instead of the salt, the game has failed. In that case, 
I can gesture in the direction of the salt and say `no, no, the {\itshape salt}
please', and then she hopefully realises and gives the salt to me. Failure 
is common in natural language dialogues and may be caused by 
many factors. For example, the salmon could indeed 
be close to the salt and my pronunciation of the word `salt'
may have sounded a bit like `salmon', perhaps because there was
loud music playing in the background or perhaps
my friend does not understand English. 
A failure is often 
an opportunity to negotiate how something will be expressed
in the future. For example, the hearer may pick up a new word
or the speaker may realise that a certain word is 
not appropriate in this particular context.

@@

The environment of the Talking Heads is 
not fixed. The human experimenter 
may change the position of objects, add new kinds of 
objects, or eliminate
others. Consequently a strategy of naming individual objects
will not work. It would lead to a proliferation of 
proper names and it would require the Talking Heads to
recognise objects, which is very difficult to do.\footnote{For a thorough exposition of the difficulties of 
object recognition, see \cite{Ullman:1996}. 
Object constancy comes fairly late in the 
acquisition of a child's ontology, as Piaget's conservation
experiments have shown. Language probably plays an important
role in forming the notion of an object.}
Indeed, humans don't exclusively use proper names in 
natural language conversations 
either. We say `could you give me the red small square' as 
opposed to `could you give me O\_143'. Natural language
words like `red' or `small' name perceptually grounded categories
and syntactic structures indicate how they should
be combined and used to find the topic. The relation between 
a language expression and its referent is therefore 
always indirect. This is summarised
in the semiotic square (\figref{triangle}), which I will use throughout this book to 
help understand and analyse the nature of language communication. \is{semiotic square}
The semiotic square relates the four entities
involved in a verbal interaction: 
\begin{itemize}
\item An {\itshape utterance}, such as `small red square', which is 
transmitted as a physical 
signal from one agent to another one through the 
external environment. It is written between double quotes. 
\item A {\itshape meaning}, which consists of categories like [\textsc{red}], [\textsc{small}], 
or combinations of categories, like \{[\textsc{red}] [\textsc{small}]\}. Labels of 
categories are written in capital letters between 
square brackets. 
\clearpage
\item An {\itshape image segment}, denoted as \emph{S143}, which is 
a segment of the image perceived through the camera.
\item A {\itshape referent}, like object \emph{O143}, which is 
an entity in the real world. 
\end{itemize}

@@

\item The speaker must then {\itshape verbalise} this conceptualisation. 
He must use his language system to find words and 
syntactic constructions expressing this meaning.
For example, he might choose `blauw' (if he speaks
Dutch), or `bleu' (if he speaks French),
to convey the category [\textsc{blue}].

@@

I deliberately left an important aspect of language out. 
In the case of physical agents, the form cannot be transmitted
directly but needs to be articulated in speech sounds, 
written signs, or gestures, to create a true utterance. This additional 
complexity will not be discussed further in this book -- even though 
it is a fascinating topic in its own right.\footnote{See for a state of the art review: \cite{Clark:1990}. 
We have already been conducting quite extensive 
research in our group on the origin of sound systems 
using similar principles as the ones discussed in this 
book for the origins of word semantics. 
See  \cite{DeBoer:1997}. The complex 
adaptive systems approach underlying this phonetics
work was already foreshadowed in work of phoneticians
like \cite{Liljencrants:1972}.}
To make a verbal interaction nevertheless complete, agents
are given a repertoire of consonants and vowels with 
which they can make random syllables and syllable combinations, 
like `wabido', `bimaku', etc. The articulation and 
recognition of these syllables is assumed to be acquired
already and transmission is engineered to be error-free. 
This way, our attention can be focused on how
ontologies, lexicons, and grammars may emerge.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap2/figs/hpos.pdf}}
\caption{Three examples of segmented images (from top to bottom). Left and right images 
show the perceived and segmented images for two agents respectively. These images are always
different because they are dependent on the position of the agent. 
The topic is indicated by a dashed bounding box in the 
image of the speaker (on the right for the first case and on the left for the others). 
Segments which are too small are ignored. The topics have all been conceptualised
as being to the right and so the same word 
`gofubo' has been used to successfully refer to them.}
\label{f:plate10}
\end{figure}

@@

The low-level visual processes outputting
values on the various sensory channels are already quite
complicated in themselves. I will discuss them further in 
\chapref{chap:3}. I will then argue that the 
agent's repertoire of visual processes does not have to be
static and programmed in advance but evolves and adapts
through a selectionist process. New processes may ``grow''
by the random combination of primitive operations and are
pruned when they fail to yield useful information or information 
which is irrelevant in the environment in which the agent 
finds himself.

@@

Here is the simplest possible instance of a
language game, based on the scene in \figref{scene1-1}. 
The speaker, {\bfshape a1}, has picked the triangle as the topic.
To give a verbal hint, he needs 
to conceptualise this topic, which means in this 
specific case, to find a 
category (or set of categories) which distinguishes the 
topic from the other objects in the context. Here 
the context contains only one additional 
object, namely the rectangle. The category
[\textsc{vpos} 0.0–0.5]$_{\mathbf a1}$ (lower) fits the criteria. 
[\textsc{vpos} 0.0–0.5] is valid when \textsc{vpos}$ < 0.5$
which is the case for the triangle but not for the rectangle. 
{\bfshape a1} has an association in his lexicon 
relating [\textsc{vpos} 0.0–0.5]$_{\mathbf  a1}$ with the word `lu', 
{\bfshape a1} retrieves this word and transmits it to 
the hearer, which is agent {\bfshape a2}.

{\bfshape a2} has stored in his lexicon an
association between `lu' and [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$, and so
hypothesises that [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$ must be the meaning of `lu'. 
When this category is applied to the present scene, in other words,
when {\bfshape a2} filters out the objects whose value for \textsc{vpos} do not 
fall in the region $[0.0–0.5]$, only one 
remaining object is obtained, the triangle. Hence {\bfshape a2} concludes that 
this must be the topic and points to it.

@@

There are (at least) two ways in which a game similar to game 125, 
but played earlier might have failed. First of all it can be that
the speaker does not have a word yet for the meaning he wants
to convey. The speaker then invokes a strategy 
to repair this failure. The simplest strategy is to 
create a new word for the present meaning. This is how 
{\bfshape a1} might have created the word `lu', and associated it 
with [\textsc{vpos} 0.0–0.5] in his lexicon. Such simple constructive steps cause
new words to enter the lexicon.

Second, it can be that the hearer does not know the word. The 
game then fails and the speaker points to 
the topic so that the hearer can make 
an educated guess what the meaning might have been:
If the hearer is able to recover a possible topic
from the non-verbal hint given by the speaker, 
he himself can seek a distinctive category or set of categories 
that delineates this topic from the other objects in 
the context just as the speaker has done. It is 
possible (although not necessary as we will see)
that the hearer {\bfshape a2} arrives at his 
own version of the same category, namely [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$. The 
hearer now stores in his lexicon a new association between the 
form heard, `lu', and the guessed meaning [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$. 
Based on this extended lexicon, he will succeed in the same game 
in the future and he can use `lu' himself to verbalise [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$
when talking to other agents. This is the way new words
spread in the population.

@@

The observant reader will have noticed immediately that 
I have swept some important difficulties under the carpet. 
First, it could very well be that unknown to the speaker 
another agent had already used the word `lu' 
for another meaning, so that there are now 
two alternative meanings for `lu' in the lexicon. `lu' 
has become ambiguous. Second, it may be the hearer {\bfshape a2} already 
had a word for [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$, for example `bomida', 
so that there are now two synonyms for [\textsc{vpos} 0.0–0.5] in the lexicon. 
Synonymy and ambiguity are very common in 
natural languages and unavoidable when a group of 
distributed autonomous agents creates a language without
a central co-ordinator. This implies that 
the agents' lexicon must be sophisticated enough 
to support multiple associations. An agent must be able to 
store the different meanings being used for the same word, and the 
different words being used for the same meaning.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap2/figs/competition-1.pdf}}
\caption{\label{competition1}The graph shows the competition between different forms for
expressing one meaning in a population of 20 agents. The graph plots 
the frequency of each form in the population, more precisely the percentage
of agents that prefer a particular word. A complex dynamic
process unfolds with periods where one word (first `xu' and then `fepi') dominates.} 
\end{figure}

@@

Another difficulty which I did not deal with when discussing Game 25 above,
is that the hearer may conceptualise
the scene differently from the speaker. For example, 
the triangular object is not only located at the lower half
of the scene and the rectangle at the upper half, 
but it is also to the left, i.e. with $\textsc{hpos} < 0.5$, whereas
the rectangle is to the right. It follows that the 
hearer could just as well have hypothesised that `lu' means [\textsc{hpos} 0.0–0.5] (left)
and not [\textsc{vpos} 0.0–0.5] (lower).

Is this bad? It depends on the situations being encountered
in the future. When the game is played again for the
same scene with {\bfshape a1} saying `lu' to mean [\textsc{vpos} 0.0–0.5], and 
{\bfshape a2} interpreting `lu' as [\textsc{hpos} 0.0–0.5], the game would succeed! 
Communicative success is achieved whenever the 
hearer recovers the referent chosen by the 
speaker; it is not required that they 
use exactly the same meaning. In fact, the hearer can never know
for sure what meaning was initially conceptualised by the speaker and 
neither can the speaker know which meaning was understood 
by the hearer because they cannot look inside each
other's head. The meaning can be quite different, 
as we have just seen, as long as it is compatible.

\clearpage
Even more remarkably, if a topic which is at the lower
portion of the scene, is always 
on the left side and vice-versa, there would always be 
success despite the different meanings of `lu', 
and the players would never discover
that each means something else by `lu'.

Similar situations arise in human natural language communication
as well, particularly for words
which are not stabilised yet or whose interpretation 
depends strongly on the non-verbal information provided
by the context. They also arise when the speaker or hearer
have different sensory modalities or sensibilities. 
For example, a colour-blind person (unable to 
make the distinction between red and green) will recognise
the red traffic light by its position. For that person, 
`stop for the red light' does not mean stop for 
the light that has the colour red but stop when the 
upper-most light is lit.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap2/figs/scene1-2.pdf}}
\caption{\label{scene1-2}A second scene which 
can be used to disambiguate `lu'.}
\end{figure}

Consider now another scene (\figref{scene1-2}). The speaker
is again {\bfshape a1} and categorises the bottom triangle as 
[\textsc{vpos} 0.0–0.5]$_{\mathbf  a1}$, meaning
the object at the lower half of the scene. Assuming that the
hearer {\bfshape a2} has associated `lu' with [\textsc{hpos} 0.0–0.5]$_{\mathbf  a2}$, he
would interpret `lu' as [\textsc{hpos} 0.0–0.5]$_{\mathbf  a2}$ (to the left). But this 
does not yield a single 
referent and therefore the game fails. 
This failure is an opportunity for the hearer to repair his 
hypotheses about the possible meanings of `lu'.
When he conceptualises the scene himself, 
he finds that [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$ distinguishes the triangle
from the circle. A new association between 
`lu' and [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$ is stored. The old association
is not removed but enters in competition with the new
one, and gradually one meaning will come to dominate
by the winner-take-all mechanism discussed earlier on.

@@

Through such disambiguating situations, meanings of words
get clarified and the lexicons of the agents
become more similar. Note that the 
dominating meaning of `lu' can still go in 
two directions. For example, {\bfshape a2}
could have used `lu' with {\bfshape a1} 
in a situation where its only possible meaning 
is [\textsc{hpos} 0.0–0.5] (left). Then {\bfshape a1}, now playing the role
of hearer, would have stored the association between
`lu' and [\textsc{hpos} 0.0–0.5]. If this happened often 
enough [\textsc{hpos} 0.0–0.5] (left) might
have become the dominant meaning of `lu', instead of 
[\textsc{vpos} 0.0–0.5] (lower). This shows clearly that the 
evolution of language will never be predictable. At a critical 
bifurcation, small preferential differences between 
the agents or the chance occurrence of certain situations
may tilt the competition in one direction or the other.\footnote{This high sensitivity to initial conditions is 
one of the characteristics of a chaotic dynamic
system, see \cite{Lorenz:1993}.
Indeed, as I will explore in more detail later, 
evolving languages have all the characteristics
of complex adaptive systems, including the 
potential for punctuated equilibria.}
There is no right
or wrong solution in the language game and no one has
any more rights than anyone else.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap2/figs/left-right.pdf}}
\caption{\label{2:left-right}Two Talking Heads are shown {\bfshape a1} and 
{\bfshape a2} each seeing the scene from a slightly different viewpoint.}
\end{figure}
When this occurs in real dialogues, the form-meaning 
pairs lexicalising the distinctions left and right destabilise,
because sometimes it gets positive feedback as 
it succeeds in the game, and at other times it 
gets negative feedback. 
This is one of the reasons why human categories are often 
relative and scaled with respect to a context.
For example, we say `the triangle to the left of the square', in 
other words left with respect to the square, to avoid the uncertainty
inherent in the absolute use of `left'. Humans scale the size 
of objects with respect to the scene, so that small 
objects are small compared to the others in the scene 
and not small in an absolute sense.

@@

Another major difference with classical theories of 
meaning and reference is 
that the repertoire of meanings and form-meaning pairs is open-ended
and subject to change at any time. Logicians would say that the 
truth conditions of the forms are non-monotonic. \is{non-monotonicity}
Non-monotonicity is unavoidable in the case of grounded 
situated agents with limited rationality.\footnote{Examples of these processes are described in: 
\cite{Bybee:1994}, \cite{Traugott:1991}.
For attempts to explain these ``grammaticalisation'' 
processes in terms of general cognitive operations, see  
\cite{Heine:1997}.}
Agents should always be allowed to 
introduce new forms or recruit existing forms to express new
meanings, simply because the set of meanings must expand
to cope with novel contexts and new communicative
situations. After enough interactions among the 
members of the same group in a relatively stable 
environment, we expect there to be
a stable set of conventionalised form-meaning 
relations, but it cannot be
expected that everything anyone ever wants to say is
already conventionalised, and so there will always
be turbulence at the fringes. The idea that the lexicon
and the syntax of a language are static
entities is completely false. Both are in constant flux.
Non-monotonicity is consequently one of the big 
topics in logic-based approaches to artificial intelligence.

@@

Each language user employs his own ideolect which is 
as well as possible co-ordinated with that of other
language users but there is never complete similarity 
and never absolute stability.  
This explains perhaps why linguists have such a hard
time to pin down ``the'' language of a community. As soon 
as someone speaks, language changes.

>>>>

Virtual agents cannot interact and 
an interaction between two \dblenquote{real} agents
can only take place when they are both physically present
in the same location. Thus an agent can travel from
Paris to Tokyo at the blink of an eye, rather than having
to take a plane, but two agents can only interact when they
are instantiated in the same physical environment. 
It is in principle possible that agents in different 
physical locations describe to each other the environment that they 
see (but which the other one does not see), just as we 
would do in a telephone conversation. This is only possible,
however, after the agents have had sufficient interactions
with each other in a shared physical world to have 
developed and learned a grounded shared language.

@@

The speaker first looks at one area of the white board and 
directs the attention of the hearer to the same area.\footnote{I will explain later how exactly agents
decide on a particular scene and how they are able to 
draw each other's attention to specific areas of the 
white board.}
The objects located in this area constitute the {\itshape context}. 
The speaker then chooses one object from the context, which 
I will call the {\itshape topic}, and gives a verbal hint
to the hearer. The {\itshape verbal hint} is an expression
that identifies the topic 
with respect to the other objects in the context. For example, 
if the context contains a red square, a blue triangle, 
and a green circle, then the speaker may say something like 
\smplenquote{the red one} to identify the red square as the topic. 
If the context contains also a red triangle, he has to be more 
precise and say something like \smplenquote{the red square} to delineate
the topic from the red triangle as well as
from the blue square. 
Of course, the Talking Heads do not say \smplenquote{the red square}
but use their own language and concepts which are not 
necessarily the same as those used in English. For example, 
they may say \smplenquote{malewina} to mean [\textsc{upper extreme}-\textsc{left low}-\textsc{redness}].

Based on the verbal hint, the hearer tries to guess what
topic the speaker has chosen, and communicates his choice 
to the speaker by pointing to the object. Given that 
the robots do not have arms, pointing is realised by 
focusing on an object. One robot can \dblenquote{see} in which direction another one 
is looking, and thus know where he is \dblenquote{pointing}. 
The game succeeds if the topic guessed by the hearer is 
equal to the topic chosen by the speaker. 
The game fails if the guess was wrong or 
if the speaker or the hearer failed at some earlier point in the 
game.

@@

The Guessing Game is one of the common things we do with language. 
For example, I play a similar game when I sit with a friend
at the dining table and say \smplenquote{could you give me the salt}. 
If she guesses correctly what I mean and hands 
me the salt, the game has succeeded. If she looks at me 
with a puzzled face (maybe she does not speak English) or hands me
the salmon instead of the salt, the game has failed. In that case, 
I can gesture in the direction of the salt and say `no, no, the {\itshape salt}
please', and then she hopefully realises and gives the salt to me. Failure 
is common in natural language dialogues and may be caused by 
many factors. For example, the salmon could indeed 
be close to the salt and my pronunciation of the word \smplenquote{salt}
may have sounded a bit like \smplenquote{salmon}, perhaps because there was
loud music playing in the background or perhaps
my friend does not understand English. 
A failure is often 
an opportunity to negotiate how something will be expressed
in the future. For example, the hearer may pick up a new word
or the speaker may realise that a certain word is 
not appropriate in this particular context.

@@

The environment of the Talking Heads is 
not fixed. The human experimenter 
may change the position of objects, add new kinds of 
objects, or eliminate
others. Consequently a strategy of naming individual objects
will not work. It would lead to a proliferation of 
proper names and it would require the Talking Heads to
recognise objects, which is very difficult to do.\footnote{For a thorough exposition of the difficulties of 
object recognition, see \cite{Ullman:1996}. 
Object constancy comes fairly late in the 
acquisition of a child's ontology, as Piaget's conservation
experiments have shown. Language probably plays an important
role in forming the notion of an object.}
Indeed, humans don't exclusively use proper names in 
natural language conversations 
either. We say \smplenquote{could you give me the red small square} as 
opposed to \smplenquote{could you give me O\_143}. Natural language
words like \smplenquote{red} or \smplenquote{small} name perceptually grounded categories
and syntactic structures indicate how they should
be combined and used to find the topic. The relation between 
a language expression and its referent is therefore 
always indirect. This is summarised
in the semiotic square (\figref{triangle}), which I will use throughout this book to 
help understand and analyse the nature of language communication. \is{semiotic square}
The semiotic square relates the four entities
involved in a verbal interaction: 
\begin{itemize}
\item An {\itshape utterance}, such as \smplenquote{small red square}, which is 
transmitted as a physical 
signal from one agent to another one through the 
external environment. It is written between double quotes. 
\item A {\itshape meaning}, which consists of categories like [\textsc{red}], [\textsc{small}], 
or combinations of categories, like \{[\textsc{red}] [\textsc{small}]\}. Labels of 
categories are written in capital letters between 
square brackets. 
\clearpage
\item An {\itshape image segment}, denoted as \emph{S143}, which is 
a segment of the image perceived through the camera.
\item A {\itshape referent}, like object \emph{O143}, which is 
an entity in the real world. 
\end{itemize}

@@

\item The speaker must then {\itshape verbalise} this conceptualisation. 
He must use his language system to find words and 
syntactic constructions expressing this meaning.
For example, he might choose \smplenquote{blauw} (if he speaks
Dutch), or \smplenquote{bleu} (if he speaks French),
to convey the category [\textsc{blue}].

@@

I deliberately left an important aspect of language out. 
In the case of physical agents, the form cannot be transmitted
directly but needs to be articulated in speech sounds, 
written signs, or gestures, to create a true utterance. This additional 
complexity will not be discussed further in this book -- even though 
it is a fascinating topic in its own right.\footnote{See for a state of the art review: \cite{Clark:1990}. 
We have already been conducting quite extensive 
research in our group on the origin of sound systems 
using similar principles as the ones discussed in this 
book for the origins of word semantics. 
See  \cite{DeBoer:1997}. The complex 
adaptive systems approach underlying this phonetics
work was already foreshadowed in work of phoneticians
like \cite{Liljencrants:1972}.}
To make a verbal interaction nevertheless complete, agents
are given a repertoire of consonants and vowels with 
which they can make random syllables and syllable combinations, 
like \smplenquote{wabido}, \smplenquote{bimaku}, etc. The articulation and 
recognition of these syllables is assumed to be acquired
already and transmission is engineered to be error-free. 
This way, our attention can be focused on how
ontologies, lexicons, and grammars may emerge.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap2/figs/hpos.pdf}}
\caption{Three examples of segmented images (from top to bottom). Left and right images 
show the perceived and segmented images for two agents respectively. These images are always
different because they are dependent on the position of the agent. 
The topic is indicated by a dashed bounding box in the 
image of the speaker (on the right for the first case and on the left for the others). 
Segments which are too small are ignored. The topics have all been conceptualised
as being to the right and so the same word 
\smplenquote{gofubo} has been used to successfully refer to them.}
\label{f:plate10}
\end{figure}

@@

The low-level visual processes outputting
values on the various sensory channels are already quite
complicated in themselves. I will discuss them further in 
\chapref{chap:3}. I will then argue that the 
agent's repertoire of visual processes does not have to be
static and programmed in advance but evolves and adapts
through a selectionist process. New processes may \dblenquote{grow}
by the random combination of primitive operations and are
pruned when they fail to yield useful information or information 
which is irrelevant in the environment in which the agent 
finds himself.

@@

Here is the simplest possible instance of a
language game, based on the scene in \figref{scene1-1}. 
The speaker, {\bfshape a1}, has picked the triangle as the topic.
To give a verbal hint, he needs 
to conceptualise this topic, which means in this 
specific case, to find a 
category (or set of categories) which distinguishes the 
topic from the other objects in the context. Here 
the context contains only one additional 
object, namely the rectangle. The category
[\textsc{vpos} 0.0–0.5]$_{\mathbf a1}$ (lower) fits the criteria. 
[\textsc{vpos} 0.0–0.5] is valid when \textsc{vpos}$ < 0.5$
which is the case for the triangle but not for the rectangle. 
{\bfshape a1} has an association in his lexicon 
relating [\textsc{vpos} 0.0–0.5]$_{\mathbf  a1}$ with the word \smplenquote{lu}, 
{\bfshape a1} retrieves this word and transmits it to 
the hearer, which is agent {\bfshape a2}.

{\bfshape a2} has stored in his lexicon an
association between \smplenquote{lu} and [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$, and so
hypothesises that [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$ must be the meaning of \smplenquote{lu}. 
When this category is applied to the present scene, in other words,
when {\bfshape a2} filters out the objects whose value for \textsc{vpos} do not 
fall in the region $[0.0–0.5]$, only one 
remaining object is obtained, the triangle. Hence {\bfshape a2} concludes that 
this must be the topic and points to it.

@@

There are (at least) two ways in which a game similar to game 125, 
but played earlier might have failed. First of all it can be that
the speaker does not have a word yet for the meaning he wants
to convey. The speaker then invokes a strategy 
to repair this failure. The simplest strategy is to 
create a new word for the present meaning. This is how 
{\bfshape a1} might have created the word \smplenquote{lu}, and associated it 
with [\textsc{vpos} 0.0–0.5] in his lexicon. Such simple constructive steps cause
new words to enter the lexicon.

Second, it can be that the hearer does not know the word. The 
game then fails and the speaker points to 
the topic so that the hearer can make 
an educated guess what the meaning might have been:
If the hearer is able to recover a possible topic
from the non-verbal hint given by the speaker, 
he himself can seek a distinctive category or set of categories 
that delineates this topic from the other objects in 
the context just as the speaker has done. It is 
possible (although not necessary as we will see)
that the hearer {\bfshape a2} arrives at his 
own version of the same category, namely [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$. The 
hearer now stores in his lexicon a new association between the 
form heard, \smplenquote{lu}, and the guessed meaning [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$. 
Based on this extended lexicon, he will succeed in the same game 
in the future and he can use \smplenquote{lu} himself to verbalise [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$
when talking to other agents. This is the way new words
spread in the population.

@@

The observant reader will have noticed immediately that 
I have swept some important difficulties under the carpet. 
First, it could very well be that unknown to the speaker 
another agent had already used the word \smplenquote{lu} 
for another meaning, so that there are now 
two alternative meanings for \smplenquote{lu} in the lexicon. \smplenquote{lu} 
has become ambiguous. Second, it may be the hearer {\bfshape a2} already 
had a word for [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$, for example \smplenquote{bomida}, 
so that there are now two synonyms for [\textsc{vpos} 0.0–0.5] in the lexicon. 
Synonymy and ambiguity are very common in 
natural languages and unavoidable when a group of 
distributed autonomous agents creates a language without
a central co-ordinator. This implies that 
the agents' lexicon must be sophisticated enough 
to support multiple associations. An agent must be able to 
store the different meanings being used for the same word, and the 
different words being used for the same meaning.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap2/figs/competition-1.pdf}}
\caption{\label{competition1}The graph shows the competition between different forms for
expressing one meaning in a population of 20 agents. The graph plots 
the frequency of each form in the population, more precisely the percentage
of agents that prefer a particular word. A complex dynamic
process unfolds with periods where one word (first \smplenquote{xu} and then \smplenquote{fepi}) dominates.} 
\end{figure}

@@

Another difficulty which I did not deal with when discussing Game 25 above,
is that the hearer may conceptualise
the scene differently from the speaker. For example, 
the triangular object is not only located at the lower half
of the scene and the rectangle at the upper half, 
but it is also to the left, i.e. with $\textsc{hpos} < 0.5$, whereas
the rectangle is to the right. It follows that the 
hearer could just as well have hypothesised that \smplenquote{lu} means [\textsc{hpos} 0.0–0.5] (left)
and not [\textsc{vpos} 0.0–0.5] (lower).

Is this bad? It depends on the situations being encountered
in the future. When the game is played again for the
same scene with {\bfshape a1} saying \smplenquote{lu} to mean [\textsc{vpos} 0.0–0.5], and 
{\bfshape a2} interpreting \smplenquote{lu} as [\textsc{hpos} 0.0–0.5], the game would succeed! 
Communicative success is achieved whenever the 
hearer recovers the referent chosen by the 
speaker; it is not required that they 
use exactly the same meaning. In fact, the hearer can never know
for sure what meaning was initially conceptualised by the speaker and 
neither can the speaker know which meaning was understood 
by the hearer because they cannot look inside each
other's head. The meaning can be quite different, 
as we have just seen, as long as it is compatible.

\clearpage
Even more remarkably, if a topic which is at the lower
portion of the scene, is always 
on the left side and vice-versa, there would always be 
success despite the different meanings of \smplenquote{lu}, 
and the players would never discover
that each means something else by \smplenquote{lu}.

Similar situations arise in human natural language communication
as well, particularly for words
which are not stabilised yet or whose interpretation 
depends strongly on the non-verbal information provided
by the context. They also arise when the speaker or hearer
have different sensory modalities or sensibilities. 
For example, a colour-blind person (unable to 
make the distinction between red and green) will recognise
the red traffic light by its position. For that person, 
\smplenquote{stop for the red light} does not mean stop for 
the light that has the colour red but stop when the 
upper-most light is lit.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap2/figs/scene1-2.pdf}}
\caption{\label{scene1-2}A second scene which 
can be used to disambiguate \smplenquote{lu}.}
\end{figure}

Consider now another scene (\figref{scene1-2}). The speaker
is again {\bfshape a1} and categorises the bottom triangle as 
[\textsc{vpos} 0.0–0.5]$_{\mathbf  a1}$, meaning
the object at the lower half of the scene. Assuming that the
hearer {\bfshape a2} has associated \smplenquote{lu} with [\textsc{hpos} 0.0–0.5]$_{\mathbf  a2}$, he
would interpret \smplenquote{lu} as [\textsc{hpos} 0.0–0.5]$_{\mathbf  a2}$ (to the left). But this 
does not yield a single 
referent and therefore the game fails. 
This failure is an opportunity for the hearer to repair his 
hypotheses about the possible meanings of \smplenquote{lu}.
When he conceptualises the scene himself, 
he finds that [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$ distinguishes the triangle
from the circle. A new association between 
\smplenquote{lu} and [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$ is stored. The old association
is not removed but enters in competition with the new
one, and gradually one meaning will come to dominate
by the winner-take-all mechanism discussed earlier on.

@@

Through such disambiguating situations, meanings of words
get clarified and the lexicons of the agents
become more similar. Note that the 
dominating meaning of \smplenquote{lu} can still go in 
two directions. For example, {\bfshape a2}
could have used \smplenquote{lu} with {\bfshape a1} 
in a situation where its only possible meaning 
is [\textsc{hpos} 0.0–0.5] (left). Then {\bfshape a1}, now playing the role
of hearer, would have stored the association between
\smplenquote{lu} and [\textsc{hpos} 0.0–0.5]. If this happened often 
enough [\textsc{hpos} 0.0–0.5] (left) might
have become the dominant meaning of \smplenquote{lu}, instead of 
[\textsc{vpos} 0.0–0.5] (lower). This shows clearly that the 
evolution of language will never be predictable. At a critical 
bifurcation, small preferential differences between 
the agents or the chance occurrence of certain situations
may tilt the competition in one direction or the other.\footnote{This high sensitivity to initial conditions is 
one of the characteristics of a chaotic dynamic
system, see \cite{Lorenz:1993}.
Indeed, as I will explore in more detail later, 
evolving languages have all the characteristics
of complex adaptive systems, including the 
potential for punctuated equilibria.}
There is no right
or wrong solution in the language game and no one has
any more rights than anyone else.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap2/figs/left-right.pdf}}
\caption{\label{2:left-right}Two Talking Heads are shown {\bfshape a1} and 
{\bfshape a2} each seeing the scene from a slightly different viewpoint.}
\end{figure}
When this occurs in real dialogues, the form-meaning 
pairs lexicalising the distinctions left and right destabilise,
because sometimes it gets positive feedback as 
it succeeds in the game, and at other times it 
gets negative feedback. 
This is one of the reasons why human categories are often 
relative and scaled with respect to a context.
For example, we say \smplenquote{the triangle to the left of the square}, in 
other words left with respect to the square, to avoid the uncertainty
inherent in the absolute use of \smplenquote{left}. Humans scale the size 
of objects with respect to the scene, so that small 
objects are small compared to the others in the scene 
and not small in an absolute sense.

@@

Another major difference with classical theories of 
meaning and reference is 
that the repertoire of meanings and form-meaning pairs is open-ended
and subject to change at any time. Logicians would say that the 
truth conditions of the forms are non-monotonic. \is{non-monotonicity}
Non-monotonicity is unavoidable in the case of grounded 
situated agents with limited rationality.\footnote{Examples of these processes are described in: 
\cite{Bybee:1994}, \cite{Traugott:1991}.
For attempts to explain these \dblenquote{grammaticalisation} 
processes in terms of general cognitive operations, see  
\cite{Heine:1997}.}
Agents should always be allowed to 
introduce new forms or recruit existing forms to express new
meanings, simply because the set of meanings must expand
to cope with novel contexts and new communicative
situations. After enough interactions among the 
members of the same group in a relatively stable 
environment, we expect there to be
a stable set of conventionalised form-meaning 
relations, but it cannot be
expected that everything anyone ever wants to say is
already conventionalised, and so there will always
be turbulence at the fringes. The idea that the lexicon
and the syntax of a language are static
entities is completely false. Both are in constant flux.
Non-monotonicity is consequently one of the big 
topics in logic-based approaches to artificial intelligence.

@@

Each language user employs his own ideolect which is 
as well as possible co-ordinated with that of other
language users but there is never complete similarity 
and never absolute stability.  
This explains perhaps why linguists have such a hard
time to pin down \dblenquote{the} language of a community. As soon 
as someone speaks, language changes.
